R Notebook

### 2021-07-30 draft-implementation-2.tex
merged
Merge branch 'main' of https://github.com/edoardo-marcelli/Progetto-Bayesian

# Conflicts:
#	prova_knit_finale.log
#	prova_knit_finale.pdf

]{book}
\usepackage{amsmath,amssymb}

@@

@@

  pdftitle={Final draft},

@@

@@

\usepackage{txfonts}
\usepackage{amsmath}
\usepackage[bindingoffset=1.5cm, left=3cm, right=3cm, top=3cm, bottom=3cm]{geometry}
\usepackage{MnSymbol} %allows to insert the QED at the end of the proofs
\usepackage{hyperref} %set the hyperlink to cite sections, equations...
\urlstyle{same}
\usepackage{comment}
\usepackage{enumitem}
\usepackage{ntheorem}
\usepackage{lipsum}
\usepackage{amssymb}
\theoremstyle{break}
\newtheorem{thm}{Theorem}% theorem counter resets every \subsection
\renewcommand{\thethm}{\arabic{thm}}
\newtheorem{proposition}{Proposition}
\newtheorem{assumption}{Assumption}
\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}
\theoremstyle{nonumberplain}
\newtheorem{proof*}{Proof.}
\newtheorem{definition}{Definition}
\renewcommand{\P}{\mathbb{P}}
\newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
\setcounter{MaxMatrixCols}{10}
\setlength{\oddsidemargin}{.0in}
\setlength{\textwidth}{6 in}
\setlength{\topmargin}{-.1in}
\setlength{\textheight}{8.1in}
\renewcommand{\baselinestretch}{1.5}
\usepackage{amssymb}
\newcommand{\E}{\mathbb E}
\newcommand{\Q}{\mathbb Q}
\newcommand{\M}{\mathbb M}
\newcommand{\R}{\mathbb R}
\newcommand{\B}{\mathcal B}
\newcommand{\X}{\mathcal X}
\newcommand{\Y}{\mathcal Y}
\newcommand{\Pd}{\mathbb P}
\newcommand{\N}{\mathcal N}

@@

\floatplacement{figure}{}

@@

\title{Final draft}
\author{}
\date{\vspace{-2.5em}}

\begin{document}
\frontmatter
\maketitle

\mainmatter
\chapter*{Introduction}

\chapter{Markov Processes and State-Space Models}
\section{Notation}

Let us fix the basic notation that will be employed henceforth. Upper
case letters, e.g.~\(X_t, Y_t,\) denote random variables, while lower
case letters denote the realizations. Finite sequences are made compact
using the semi-colon notation, e.g.~\(x_{0:t}=(x_0, x_1, \dots, x_t)\)
for some \(t \geq 0.\) The sets of outcomes are denoted by upper case
calligraphic letters, e.g.~\(\mathcal{X}, \mathcal{Y}.\) The
\(\sigma\)-algebra of a set \(\mathcal{X}\) reads
\(\mathcal{B}(\mathcal{X}).\) The pair
\((\mathcal{X}, \mathcal{B}(\mathcal{X}))\) is a measurable space. The
densities are represented with lower case cursive letters. For example,
the density of \(X_t\) in \(x_t\) is \(p_t^\theta(x_t),\) where
\(\theta\in\Theta\) is some (eventually known) vector, while the density
of \(X_t\) in \(x_t,\) given that \(X_{t-1}\) has previously taken value
\(x_{t-1},\) is \(p_t^\theta(x_t\mid x_{t-1}).\) Note that
\(p_t^\theta(x_t\mid x_{t-1})\) is a member of the parametric family
with parameters \((\theta, x_{t-1})\). The probability distribution of a
random variable \(X_t\) is represented by \(\mathbb{P}_t(dx_t),\) while
the probability distribution of a sequence \(X_{0:t}\) is represented by
\(\mathbb{P}_t(dx_{0:t}).\) The Dirac measure, i.e., the measure that
assigns probability \(1\) to the singleton \(\{x\},\) is denoted by
\(\delta_x(dy)\).The expectation operator is denoted by \(\mathbb{E},\)
where the section of \(\mathbb{E}\) at \(\mathbb{Q},\) i.e.,
\(\mathbb{E}_{\mathbb{Q}}\) denotes the operator computed with respect
to the probability distribution \(\mathbb{Q}.\) Let \(\varphi\) be a
function defined on the random variable \(X\propto\mathbb{Q}.\) Then, we
denote the expected value of \(\varphi(X)\) with \[
    \mathbb{E}_{\mathbb{Q}}[\varphi(X)]=
    \int_{\mathcal{X}}\varphi(x)\mathbb{Q}(dx).
\] Convergence in distribution is denoted with the symbol \(\implies.\)
So, \(X\implies\mathcal N(0,1)\) means that the random variable \(X\)
converges in distribution to a normal Gaussian.

\section{Markov Processes}

Let \((\mathcal{X}, \mathcal{B}(\mathcal{X}))\) and
\((\mathcal{Y}, \mathcal{B}(\mathcal{Y}))\) be two (eventually equal)
measurable spaces.

\begin{definition}
A function $P(x,dy): (\mathcal{X}, \mathcal{B}(\mathcal{X}))\rightarrow [0,1]$ such that 
\begin{itemize}
    \item for all $x\in\mathcal{X}, P(x,\cdot)$ is a probability measure on $(\mathcal{Y}, \mathcal{B}(\mathcal{Y})$, and 
    \item for all subsets $A\in\mathcal{B}(\mathcal{Y}),$ the map $x\mapsto P(x,A)$ is measurable in $\mathcal{B}(\mathcal{X})),$
\end{itemize}
is defined a probability kernel from $(\mathcal{X}, \mathcal{B}(\mathcal{X}))$ to $(\mathcal{Y}, \mathcal{B}(\mathcal{Y})).$
\end{definition}

Notice that for any random variables \(X_1\) and \(X_0=x_0,\) \[
    \mathbb{P}_1(dx_{0:1})=\mathbb{P}_0(dx_{0})P_1(x_0,dx_1),
\] where \(P_1(x_0,dx_1)\) is a probability kernel from
\((\mathcal{X}, \mathcal{B}(\mathcal{X}))\) to
\((\mathcal{X}, \mathcal{B}(\mathcal{X})).\) This remark, provides some
intuition about the link between probability kernels and conditional
probabilities. As a matter of fact, it is possible to show that for
given \(X_{0:1}\) with joint probability distribution
\(\mathbb{P}_1(dx_{0:1}),\)
\begin{equation}\label{prob_kern_condit_prob}
    \mathbb{P}_1(X_1\in dx_1\mid X_0=x_0)=P_1(x_0,dx_1).
\end{equation}

Probability kernels can be used to define Markov processes. Before
moving to the definition, we will introduce some important concepts.
Recall that given two measures \(\mathbb{M}\) and \(\mathbb{Q}\) defined
on a measurable space \((\mathcal{X}, \mathcal{B}(\mathcal{X})),\)
\(\mathbb{Q}\) is absolutely continuous with respect to \(\mathbb{M},\)
that is \(\mathbb{Q}\ll\mathbb{M},\) if for all sets
\(A\in\mathcal{B}(\mathcal{X})\) such that \(\mathbb{M}(A)=0,\) then it
must also hold that \(\mathbb{Q}(A)=0.\) If
\(\mathbb{Q}(A)/\mathbb{M}(A)\) is well-defined for all \(A,\) then the
Radon-Nikodym theorem guarantees that there exists a measurable function
\(w(x)\geq0\) such that \[
    w(x)=\frac{\mathbb{Q}(dx)}{\mathbb{M}(dx)}, 
\] meaning that, \[
    \mathbb{Q}(A)=\int_A w(x) \mathbb{M}(dx).
\]

Let \(P_1, P_2, \dots, P_T\) be a finite sequence of probability kernels
from \((\mathcal{X}, \mathcal{B}(\mathcal{X}))\) to
\((\mathcal{X}, \mathcal{B}(\mathcal{X})).\) Fix some
\(\mathbb{P}_0(dx)\) on \((\mathcal{X}, \mathcal{B}(\mathcal{X})).\)

\begin{definition}
A discrete-time Markov process is a sequence $X_{0:T}$ of random variables whose joint distribution can be written as
\begin{equation}
    \mathbb{P}_T(X_{0:T}\in dx_{0:T})=\mathbb{P}_0(dx_0)\prod\limits_{s=1}^T P_s(x_{s-1},dx_s).
\end{equation}
The set $\mathcal{X}$ is the state-space, $\mathbb{P}_0$ is the initial distribution and the probability kernel $P_t$ is the transition kernel.
\end{definition}

It is possible to show that for any \(t\in T,\) \[
\mathbb{P}_T(X_t\in dx_t\mid X_{0:t-1}\in dx_{0:t-1})=
\mathbb{P}_T(X_t\in dx_t\mid X_{t-1}=x_{t-1})=\mathbb{P}_t(x_{t-1},dx_t).
\] The first equality implies conditional independence. The second
equality implies that it is possible to identify conditional
distributions with transition kernels. Assume \(\mathbb{P}_t,\) with
\(t\leq T,\) is a sequence of probability measures. Then, the following
proposition holds.

\begin{proposition}
For all $t\leq T,$ 
$$
\mathbb{P}_T(dx_{0:t})=\mathbb{P}_0(dx_0)\prod\limits_{s=1}^t P_s(x_{s-1},dx_s)=\mathbb{P}_t(dx_{0:t}).
$$
\end{proposition}

This means that the marginal distribution of \(X_{0:t}\) with respect to
\(\mathbb{P}_T\) is \(\mathbb{P}_t.\)

\section{State-Space Models}

We now define \textit{State-Space models} as Markov processes that are
partially observed. Let \(X_t\in\mathcal X\) and \(Y_t\in\mathcal Y\)
and let \(\{(X_t,Y_t)\}\) be a stochastic process. To help intuition,
suppose that \(\{Y_t\}\) is observable, i.e., coincides with a sequence
of observations, while \(\{X_t\}\) is latent. Consider the initial
distribution \(\mathbb{P}_0(dx_0)\) and the probability kernels
\(P_t(x_{t-1},dx_t)\) and \(F_t(x_t,d_t)\), where \(t=1,2,\dots\).

\begin{definition}\label{state-space-def}
The process $\{(X_t,Y_t)\}$ is a State-Space model if the joint distribution of $(X_{0:T},Y_{0:T})$ is 
\begin{align*}
  \mathbb{P}_T(X_{0:T}\in dx_{0:T},Y_{0:T}\in dy_{0:T}) &= \mathbb{P}_0(dx_0)\prod\limits_{t=1}^T P_t(x_{t-1},dx_t)\prod\limits_{t=1}^T F_t(x_t,dy_t)\\
  &= [\mathbb{P}_0(dx_0)F_0(x_0,y_0)]\prod\limits_{t=1}^T[P_t(x_{t-1},dx_t)F_t(x_t,dy_t)] \\
  &= \mathbb{P}_T(dx_{0:T})\prod \limits_{t=1}^T F_t(x_t,dy_t).
\end{align*}
\end{definition}

From the second inequality, it is apparent that \(\{(X_t,Y_t)\}\) is a
Markov process with initial distribution
\([\mathbb{P}_0(dx_0)F_0(x_0,y_0)]\) and transition kernel
\(P_t(x_{t-1},dx_t)F_t(x_t,dy_t).\) The third equality, instead, shows
that, \(\{X_t\}\) has a marginal distribution
\[\mathbb{P}_T(dx_{0:T})=\mathbb{P}_0(dx_0)\prod_{t=1}^T P_t(x_{t-1},dx_t),\]
meaning that \(\{X_t\}\) is marginally distributed as a Markov process
with initial distribution \(\mathbb{P}_T(dx_{0:T})\) and transition
kernel \(P_t(x_{t-1},dx_t).\)

Assume now that there exists a dominating measures \(\nu(dy)\) such that
\(F_t(x_t,dy_t)=f_t(y_t\mid x_t)\nu(dy_t)\) for all \(t.\) Then by the
last equality of definition \ref{state-space-def}, the joint
distribution of \((X_{0:T},Y_{0:T})\) is
\begin{equation}\label{state-space-densities}
   \mathbb{P}_T(X_{0:T}\in dx_{0:T},Y_{0:T}\in dy_{0:T})=\mathbb{P}_T(dx_{0:T})\prod\limits_{t=0}^T f_t(y_t\mid x_t)\prod\limits_{t=0}^T\vu(dy_t).
\end{equation} Integrating away \(X_{0:T},\) one gets the marginal
density \begin{equation}\label{state-space-marginal-y}
\begin{align*}
  \mathbb{P}_T(dy_{0:T}) &= \mathbb E_{\mathbb{P}_{t}}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big] \prod\limits_{s=0}^t\nu(dy_s) \\
                &= p_t(y_{0:t})\prod\limits_{s=0}^t\nu(dy_s),
\end{align*}                
\end{equation} where \(p_t(y_{0:t})\) is the likelihood function.
Dividing the left-hand side of equation \ref{state-space-densities} by
the left-hand side of equation \ref{state-space-marginal-y} and the
right-hand side of equation \ref{state-space-densities} by the
right-hand side of equation \ref{state-space-marginal-y}, one obtains
the filtering distribution \begin{equation}
\mathbb{P}_t(X_{t}\in dx_{t}\mid Y_{0:t}=y_{0:t})=\frac{1}{p_t(y_{0:t})}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big]\mathbb{P}_t(dx_{0:t}).
\end{equation}

\chapter{Linear-Gaussian case}

\textit{Linear Gaussian models} are a class of popular state-space
models. A notable property of such models is that the filtering and
predictive distributions are analytically tractable, which makes such
models very convenient in applications.

Suppose that \(\mathcal X=\mathbb R^{d_x}\) and
\(\mathcal Y=\mathbb R^{d_y}.\) Let
\(U_t\thicksim\mathcal N(0,\Sigma_t)\) and
\(V_t\thicksim\mathcal N(0,R_t)\) be two independent white-noise errors
and assume the initial distribution
\(X_0\thicksim\mathcal N(0,\Sigma_0).\) We have a linear Gaussian model
if \begin{align}
X_t &= A_tX_{t-1}+U_t \\
Y_t &= B_tX_t + V_t.
\end{align} Note that \(X_t\mid X_{t-1}\) and \(Y_t\mid X_t\) are both
normally distributed.

\begin{lemma}\label{kalman-filter}
  Let $X\thicksim\mathcal N(m_0,Q_0)$ and $Y\mid X\thicksim\mathcal N(BX,R)$, where $R$ and $Q_0$ are positive semi-definite matrices. 
  Then, it holds that
  \begin{align}
    X\mid Y=y&\thicksim\mathcal N(m_1,Q_1),\\
    Q_1&= Q_0 (I_{d_x} - B'(B Q_0 B'+ R)^{-1} B Q_0) \\
    m_1&=[I_{d_x} - Q_0 B'(B Q_0 B'+ R)^{-1} B ]m_0 + Q_0 B'(B Q_0 B'+ R)^{-1}y.
  \end{align}
\end{lemma}
\begin{proof*}
   The lemma can be proved using Bayes' Theorem. Suppose $X\thicksim\mathcal N(m_0, Q_0).$ It is easy to see that
\begin{equation} \label{normal-distrib-rewrite}
        p(x)\propto \text{exp}\Big(-\frac{1}{2}X'AX + X'B\Big),
\end{equation}
with $A=Q_0^{-1}$ and $B=Am_0.$
Now, by Bayes' Theorem,

    \begin{align*}
        p(x\mid Y=y)  &\propto p(x) p(y\mid x) \\
                    &\propto \text{exp}\Big(-\frac{1}{2}X' Q_0^{-1}X + X' Q_0^{-1}m_0-\frac{1}{2}(y-BX)'R^{-1}(y-BX)) \\
                    &\propto \text{exp}(-\frac{1}{2}X' Q_0^{-1}X + X' Q_0^{-1}m_0-\frac{1}{2}X' B' R^{-1}BX+X' B' R^{-1}y)\\
                    &=\text{exp}(-\frac{1}{2}X'(Q_0^{-1}+B' R^{-1}B)X+
                    X'(Q_0^{-1}m_0+B' R^{-1}y)\Big).
    \end{align*}

Hence, it follows that 
\begin{equation*}
    Q_1=Var(X\mid Y=y)= (Q_0^{-1}+ B' R^{-1} B)^{-1}=Q_0-Q_0 B'(B Q_0 B'+R)^{-1}B Q_0
\end{equation*}
and 
    \begin{align*}
        m_1 &=\mathbb E(X\mid Y=y) = Q_1^{-1}(Q_0^{-1}m_0+B' R^{-1}y)\\
            &= (Q_0-Q_0B'(B Q_0B'+R)^{-1}B Q_0)(Q_0^{-1}m_0+B' R^{-1}y)\\
            &= m_0-Q_0B'(B Q_0B'+R)^{-1}B m_0 + Q_0 B' R^{-1}y - Q_0B'(B Q_0B'+R)^{-1}B Q_0 B' R^{-1}y \\
            &=[I_{d_x} - Q_0 B'(B Q_0 B'+ R)^{-1} B ]m_0 + Q_0 B'(B Q_0 B'+ R)^{-1}y.
    \end{align*}
\end{proof*}

Assume that
\(\mathbb{P}_{t-1}(dx_{t-1}\mid Y_{0:t-1}=y_{0:t-1})=\mathcal N(m_{t-1},Q_{t-1}).\)
Then, using the transition equation of the latent variable, by the
linearity of the Gaussian distribution, we have that the predictive
distribution is \begin{equation}
\mathbb{P}_{t-1}(dx_{t}\mid Y_{0:t-1}=y_{0:t-1})=\mathcal N(A_tm_{t-1},A_tQ_{t-1}A_t'+\Sigma_t).
\end{equation} Letting \(E_T=A_tQ_{t-1}A_t'+\Sigma_t,\) and applying
lemma \ref{kalman-filter}, it follows that the filtering distribution is
\begin{align*}
  \mathbb{P}_{t}(X_t\in dx_{t}\mid Y_{0:t-1}&=y_{0:t-1})=\mathcal N(m_t,Q_t), \text{where} \\ Q_t&=E_t[I_{d_x} - B_t'(B_tE_tB_t'+R_t)^{-1} B_tE_t] \\
  m_t &= [I_{d_x} - E_tB_t'(B_tE_tB_t'+R_t)^{-1} B_t]A_tm_{t-1}+E_tB_t'(B_tE_tB_t'+R_t)^{-1}y_t.
\end{align*}

In this section we present a practical illustration of the algorithms
discussed. For the sake of simplicity we use a random walk plus noise
model, i.e.~the most basic form of a linear Gaussian state-space
model.\\
\begin{align}
y_{t}|x_{t} & \sim N(x_{t},\sigma^{2}) \\
x_{t}|x_{t-1} & \sim N(x_{t-1},\tau^{2}) \\
x_{0} & \sim N(m_{0},C_{0})
\end{align} As already mentioned before, in this case the filtering
distribution can be computed in closed form solutions using the Kalman
filter. However, this toy example will be used also to illustrate more
involved filtering strategies described in this work. We believe indeed
that it represents a useful starting point to understand the logic of
the algorithms which may be eventually replicated when dealing with more
complex models.\\
The filtering strategy is applied to 50 simulated data. Figure XX shows
the simulated true states sequence assuming as data generating process
the Equation (2) with \(\tau^{2}=1\) and simulated observed sequence
process form Equation (1) with \(\sigma^{2}=1\).

\begin{figure}[H]

{\centering \includegraphics{final-draft_files/figure-latex/unnamed-chunk-5-1}

@@

\caption{Simulated random walk plus noise model}\label{fig:unnamed-chunk-5}
\end{figure}

@@

Our \texttt{DLM} function implement in R replicate this steps.\\

\hrule
\hrule
\texttt{DLM(data,sig2,tau2,m0,C0)}
\hrule

@@

\begin{figure}[H]

{\centering \includegraphics{final-draft_files/figure-latex/unnamed-chunk-7-1}

@@

\caption{Kalman Filtered States with credible interval (in red)}\label{fig:unnamed-chunk-7}
\end{figure}

@@

\begin{figure}[H]

{\centering \includegraphics{final-draft_files/figure-latex/unnamed-chunk-8-1}

\caption{Simulated random walk plus noise model}\label{fig:unnamed-chunk-8}
\end{figure}

\chapter{Particle Filter}

In the previous chapter we explained how the Kalman Filter is applied to
the case of a linear model with Gaussian errors to find closed-form
solution for the filtering and the predictive distributions.
Nonetheless, the majority of time-series models have a non-linear
structure and errors might follow non-Gaussian distributions.\\
In this chapter we describe \textit{particle filter} as an algorithm
that allows to update sequentially the filtering and the predictive
distributions in the non-linear and non-Gaussian case. Additionally, the
Kalman Filter can be seen as a special case of the more general particle
filter.\\
In section \ref{pf_intro} we introduce the fundamental notions of
\textit{importance sampling}, \textit{sequential importance sampling}
and \textit{resampling}. Then, in section \ref{pf_main} we describe
three particle filtering algorithms: the \textit{bootstrap}, the
\textit{guided} and the \{auxiliary particle filters\}. Instead, in
section \ref{pf_lui_west} we describe the \{Liu and West filter\}, when
particle filtering is used to estimate model's parameters. Finally, in
section \ref{pf_converg} we establish some general convergence results
for Particle Filter - Monte Carlo estimators.

\section{Introductory notions: sampling and resampling}\label{pf_intro}

Let \(\varphi: \mathcal X\rightarrow \mathbb R\) be a measurable,
integrable and bounded real-valued function, where
\((\mathcal X,\mathcal B(\mathcal X),\mathbb Q)\) is a probability
space.\footnote{$\mathcal B(\mathcal X)$ is the Borel sigma-algebra of $\mathcal X$.}
Suppose that the probability measure \(\mathbb Q\) admits a density
\(q(x)\) with respect to some measure \(\nu(dx)\) and it is feasible to
sample from that density.\\
To estimate the expected value of \(\varphi\) with respect to the
density \(q\), the basic Monte Carlo (MC) method would require to sample
\(N\) times from the target density \(q\) and to compute the average of
the function \(\varphi\) values. \(\hat \varphi_{MC}\) is the standard
MC estimator: \begin{equation*}
   \hat \varphi_{MC}=\frac{1}{N} \sum_{n=1}^N\varphi (X^n)\approx \mathbb E_q(\varphi)~~~X^n\thicksim q(\cdot)
\end{equation*} However, in many circumstances either it is impossible
to sample from the target density or there exist another density - call
it \(m(\cdot)\) - different from \(q\) that produces
\texttt{more\ efficient"\ estimates.\ \ In\ all\ the\ cases\ for\ which\ sampling\ takes\ place\ from\ a\ distribution\ different\ from\ the}correct"
(i.e., the true) one, importance sampling and resampling methods can be
applied (as augmented versions of the standard MC algorithm).

\subsection{Importance sampling}

Let us consider two probability spaces:
\((\mathcal X,\mathcal B(\mathcal X),\mathbb Q)\) and
\((\mathcal X,\mathcal B(\mathcal X),\mathbb M)\) and let the measurable
function \(w:\mathcal X\rightarrow \mathbb R\) be proportional to the
Radom-Nikodym derivative between \(\mathbb Q\) and \(\mathbb M\), i.e.,
\(w(x)\propto\frac{\mathbb Q(dx)}{\mathbb M(dx)}\).\\
The following result represents the starting point for the idea of
importance sampling.

\begin{lemma}\label{lemma_imp_sam_1}
For any measurable, integrable and bounded function $\varphi: \mathcal X\rightarrow \mathbb R$, the following holds:
\begin{equation}
    \mathbb E_\mathbb M(\varphi \cdot w)=\mathbb E_\mathbb Q(\varphi)\cdot \mathbb E_\mathbb M(w)
\end{equation}
\end{lemma}

\begin{proof*}
Note that $\mathbb Q(dx)=\frac{w(x)}{\mathbb E_\mathbb M(w)}\mathbb M(dx)$. Hence:
\begin{equation*}
    \mathbb E_\mathbb Q(\varphi)=\int_\mathcal X\varphi(x)\mathbb Q(dx)=\int_\mathcal X\varphi(x)\frac{w(x)}{\mathbb E_\mathbb M(w)}\mathbb M(dx)=\frac{\mathbb E_\mathbb M(\varphi \cdot w)}{\mathbb E_\mathbb M(w)}
\end{equation*}
$\hfill\blacksquare$
\end{proof*}

Lemma \ref{lemma_imp_sam_1} represents the key theoretical foundation of
\textbf{importance sampling} (IS), a numerical algorithm that allows to
sample from a \textit{proposal density} to estimate moments that are
computed with respect to the \textit{target density}, by means of
re-weighting the values of the target function \(\varphi\).\\
Adopting the previous notation, the target measure is \(\mathbb Q\),
while the proposal is \(\mathbb M\). IS requires to sample \(N\) values
from \(\mathbb M\), then to find the associated
\textit{normalized}\footnote{Weights are normalized when $\mathbb E_\mathbb M(W)=1$. In this case $W$ is the Radom-Nikodym derivative between $\mathbb Q$ and $\mathbb M$.}
weights \(W(X^n)\), finally to compute a weighted average of the values
taken by the function \(\varphi\). The IS is implemented according to
the following algorithm.

\begin{enumerate}
    \item Sample $N$ values from the proposal measure: $X^n\thicksim \mathbb M$ for $n=1,...,N$.
    \item Compute the associated normalized weights: $W(X^n)$ for $n=1,...,N$
    \item Compute the estimator:
    \begin{equation*}
        \hat \varphi= \frac{1}{N} \sum_{n=1}^N\varphi (X^n)W(X^n).
    \end{equation*}
\end{enumerate}

Note that \(\hat \varphi\) is an unbiased estimator for the target of
inference \(\mathbb E_\mathbb Q(\varphi)\). In fact: \begin{equation*}
    \begin{split}
        \mathbb E_\mathbb M(\hat \varphi)=\frac{1}{N} \sum_{n=1}^N \mathbb E_\mathbb M(\varphi \cdot W)=\mathbb E_\mathbb M(\varphi \cdot W)=\mathbb E_\mathbb Q(\varphi)
    \end{split}
\end{equation*} where the last equality follows from lemma
\ref{lemma_imp_sam_1} and the fact that weights are normalized to 1.

\subsubsection{Auto-normalized IS}

Usually, the Radom-Nikodym derivative is known up to a normalizing
constant and the IS algorithm has to be described in more general terms.
In the previous section the \textit{normalized} weighting function
\(W(\cdot)\) was the Radom-Nikodym derivative of \(\mathbb Q\) with
respect to \(\mathbb M\). More generally, consider the function
\(w(x)\propto\frac{\mathbb Q(dx)}{\mathbb M(dx)}\), denoted as
\textit{importance function}, proportional to the derivative
\(W(\cdot)\).\\
In this case, weights have to be \textit{auto}-normalized to obtain the
IS - Monte Carlo estimator. The auto-normalized IS is implemented
according to the following procedure.

\begin{enumerate}
    \item Sample $N$ values from the proposal measure: $X^n\thicksim \mathbb M$ for $n=1,...,N$.
    \item Compute the associated \textit{un-normalized} weights: $w(X^n)$ for $n=1,...,N$.
    \item Normalize the weights, for $n=1,...,N$:
    \begin{equation*}
        W^n:=\frac{w(X^n)}{\sum_{m=1}^Nw(X^m)}
    \end{equation*}
    \item Compute the estimator:
    \begin{equation*}
        \hat \varphi_{AN}=\sum_{n=1}^N\varphi (X^n)W^n
    \end{equation*}
\end{enumerate}

While the auto-normalized estimator is biased (since \(W^n\) involves a
ratio of random variables), it is consistent (see \ref{pf_is_conv} for
details). ~\\
It is useful to interpret IS as an algorithm that allows to find an
approximation for the probability measure \(\mathbb Q\) as a weighted
average of Dirac measures for the sampled values (\(X^n\)):
\(\delta_{X^n}(dx)\) where the weights are the auto-normalized ones
(\(W^n\)). In particular: \begin{equation}
    \mathbb Q^N(dx):=\sum_{n=1}^NW^n\delta_{X^n}(dx), ~~~X^n\thicksim \mathbb M
\end{equation} Furthermore, the IS estimator is the expected value of
the function \(\varphi\) with respect to the approximating probability
measure \(\mathbb Q^N\): \begin{equation*}
    \begin{split}
        \mathbb E_{\mathbb Q^N}(\varphi)=&\int_\mathcal X\varphi(x)\mathbb Q^N(dx)=\int_\mathcal X\varphi(x)\sum_{n=1}^NW^n\delta_{X^n}(dx)=\\
       = &\sum_{n=1}^N\int_\mathcal X\varphi(x)W^n\delta_{X^n}(dx)=\sum_{n=1}^N\varphi (X^n)W^n=\hat \varphi_{AN}
    \end{split}
\end{equation*}

\subsubsection{Convergence of the IS estimator}\label{pf_is_conv}

Consider the auto-normalized IS estimator
\(\hat \varphi= \sum_{n=1}^N\varphi (X^n)W^n\). We can write:
\begin{equation*}
   \hat \varphi_{AN} - \mathbb E_{\mathbb Q}(\varphi)=\mathbb E_{\mathbb Q^N}(\varphi)-\mathbb E_{\mathbb Q}(\varphi)=\frac{N^{-1}\sum_{n=1}^Nw(X^n)[\varphi(X^n)-\mathbb E_{\mathbb Q}(\varphi)]}{N^{-1}\sum_{n=1}^Nw(X^n)}
\end{equation*} Let
\(\bar \varphi(X):=\varphi(X)-\mathbb E_{\mathbb Q}(\varphi)\). Applying
the CLT to the numerator, assuming that
\(\mathbb E_{\mathbb M}(w^2\bar\varphi^2)<+\infty\), we get:
\begin{equation}
    \sqrt{N}\Bigg\{\frac{1}{N}\sum_{n=1}^Nw(X^n)\bar \varphi (X^n)\Bigg\}\implies \mathcal N\big(0,\mathbb E_{\mathbb M}(w^2\bar\varphi^2)\big)\label{is_conv_clt}
\end{equation} where \(\implies\) denotes convergence in distribution.
Instead, for the denominator the strong LLN implies that:
\begin{equation}
    \frac{1}{N}\sum_{n=1}^Nw(X^n)\rightarrow \mathbb E_{\mathbb M}(w) ~~~\text{ a.s.}\label{is_conv_lln}
\end{equation} Applying Slutsky's theorem, together with
\eqref{is_conv_clt} and \eqref{is_conv_lln}, we obtain: \begin{equation}
   \sqrt{N}\big[  \hat \varphi_{AN} - \mathbb E_{\mathbb Q}(\varphi)\big] \implies \mathcal N\Bigg(0,\frac{\mathbb E_{\mathbb M}(w^2\bar\varphi^2)}{[\mathbb E_{\mathbb M}(w)]^2}\Bigg)
\end{equation} Therefore, the IS (auto-normalized) estimator for
\(\varphi\) is both consistent and asymptotically normal.

\subsubsection{Effective Sample Size}

A common measure of efficiency for IS is the
\textit{effective sample size} (ESS), defined as: \begin{equation*}
    ESS(W^{1:N}):=\frac{1}{\sum_{n=1}^N(W^n)^2}=\frac{[\sum_{n=1}^N w(X^n)]^2}{\sum_{n=1}^N(w(X^n))^2}
\end{equation*} Observe that the \(ESS\) lies in the interval \([1,N]\).
For instance, if \(W^n=1/N\), as in the case of the standard MC method,
\(ESS=N\).\\
It is possible to relate the \(ESS\) to the variance of the importance
function. In fact: \begin{equation*}
    \frac{N}{ESS}=1+\frac{N^{-1}\sum_{n=1}^N(w(X^n))^2-\big[N^{-1}\sum_{n=1}^N w(X^n)\big]^2}{\big[N^{-1}\sum_{n=1}^N w(X^n)\big]^2}
\end{equation*} The second term in the numerator of the previous formula
is the square of the coefficient of variation (i.e., the ratio between
the variance and the squared mean) of the un-normalized weights. It is
clear that a lower \(ESS\) is associated with a higher variance for the
weights.

\subsection{Sequential Importance Sampling}

When dealing with state-space models, the importance sampling algorithm
has to be applied dynamically and the importance weights are updated
period-by-period. To illustrate the relevance of the
\textit{sequential importance sampling} (SIS) algorithm, we consider a
two-period stochastic process.~\\
Let us consider the approximation of the measure \(\mathbb Q_0\)
obtained at time 0 through IS from the measure \(\mathbb M_0\):
\begin{equation*}
    \mathbb Q_0^N(dx_0)=\sum_{n=1}^NW_0^n\delta_{X_0^n}(dx_0),~~\text{with } X_0^n\thicksim \mathbb M_0,~~\text{and }  W_0^n:=\frac{w_0(X_0^n)}{\sum_{m=1}^Nw_0(X_0^m)}.
\end{equation*} Moreover, let \(M_1(x_0,dx_1)\) be the transition
kernel. We want to update, sequentially, the approximating measure
\(\mathbb Q_0\) to obtain an approximation of the measure:
\begin{equation*}
    \mathbb Q_{1}(dx_{0:1})=\mathbb Q_0(dx_0)M_1(x_0,dx_1)
\end{equation*} SIS works according to the following algorithm.

\begin{enumerate}
    \item Sample $N$ values for $X_0$ from the proposal measure $\mathbb M_0$: $X_0^n\thicksim \mathbb M_0$ for $n=1,...,N$.
    \item Sample $N$ values for $X_1$ from the transition kernel $M_1$: $X_1^n\thicksim M_1(X_0^n,dx_1)$ for $n=1,...,N$.
    \item Compute the auto-normalized weights. In this case, they are computed taking into account only time 0 sampled observations since $X_1^n$ is sampled from the correct distribution: $W_1^n=W_0^n$.
\end{enumerate}

Therefore, the particle approximation of \(\mathbb Q_{1}(dx_{0:1})\) is:
\begin{equation*}
      \mathbb Q_{1}^N(dx_{0:1})=\sum_{n=1}^NW_0^n\delta_{X_0^n}(dx_0)M_1(X_0^n,dx_1)
\end{equation*}

\subsection{Sequential Importance Resampling}

A second approach can be used when sampling has to be done sequentially:
\textit{sequential importance resampling} (SIR). Differently from SIS,
an intermediate resampling stage takes place, in which time 0 values are
sampled with replacement according to a given resampling strategy
(usually, sampling indexes from a multinomial distribution with
probabilities equal to the weights computed through IS at time 0).\\
While in the SIS all the particles (i.e., the sampled values)
\texttt{survive",\ with\ the}risk" that particles associated with small
weights are carried-over (in the sense that sampling in period 1 occurs
from the kernel \(M_1(X_0^n,dx_1)\)), the same does not occur in SIR. In
this second algorithm, particles with larger weights are more likely to
be resampled and be used to ``generate" the particles in the next
period.\\
We describe the SIR more formally, focusing on the previous two-period
case.

>>>>

]{article}

@@

\usepackage{amssymb,amsmath}

@@

  pdftitle={R Notebook},

@@

\usepackage[margin=1in]{geometry}

@@

\newcommand{\R}{\textnormal{\sffamily\bfseries R}}

@@

@@

\title{R Notebook}
\author{}
\date{\vspace{-2.5em}}

\begin{document}
\maketitle

\begin{figure}[ht]

{\centering \includegraphics{draft-implementation-2_files/figure-latex/unnamed-chunk-3-1}

@@

\caption{Simulated random walk plus noise model}\label{fig:unnamed-chunk-3}
\end{figure}

@@

Our \texttt{DLM} function implement in
\textnormal{\sffamily\bfseries R} replicate this steps.\\

\hfill\break
\texttt{DLM(data,sig2,tau2,m0,C0)}\\

@@

\begin{figure}[ht]

{\centering \includegraphics{draft-implementation-2_files/figure-latex/unnamed-chunk-5-1}

@@

\caption{Kalman Filtered States with credible interval (in red)}\label{fig:unnamed-chunk-5}
\end{figure}

@@
